---
title: "Homework 1 - part 3"
author: "Francesco MONTI"
date: "2022-11-17"
output:
  pdf_document: default
  html_document:
    df_print: paged
option: warning = F
editor_options:
  chunk_output_type: console
---

```{r libraries, echo=F}
library(stringr)
library(prettyR)
library(ggplot2)
library(tidyverse)
library(ggthemes)
library(colorRamps)
library(psych)
library(Hmisc)
library(knitr)

```

Nota bene: output will go to the console, as defined in the global options.

## 1. Import the file "BDD_VICAN.csv"

```{r}
data <- read.csv("BDD_VICAN.csv", sep = ";", dec = ",", encoding = "UTF-8")

names(data) = tolower(names(data)) # removing capital letters as working with them can be annoying
```

## 2. Display the first lines of the dataset. Display the lines 1; 4; 18; 103 of the dataset

```{r echo=F}
head(data) # first lines

data[c(1, 4, 18, 103), ] # lines 1,4,18,103

```

## 3. How many variables and observations are there?

```{r echo=F}
ncol(data) # variables count

nrow(data) # observations count
```

## 4. Does this file contain any missing values?

```{r}
sapply(data, function(x) sum(is.na(x))) # NAs by variable

sum(is.na(data)) # global NAs
```

On a first impression, the dataframe looks to be free of any NAs. We'll see later that this is not true: the variable "q5_eortc_fatigue_r1" as been incorrectly identified as "character" as missing values have been tagged as "!NULL" rather than leaving the cells empty.

As a sidenote, there is no description of "q5_eortc_fatigue_r1" in the statement of the homework

## 5. What is the nature of the variables studied?

```{r}
str(data)
```

## 6. Some of the variables are in the wrong format, for example, a qualitative variable in "numeric" format. Based on the description of each variable (found at the beginning of this exercise), re-code the variable(s) into the correct format

```{r}
# replacing "," with "." is necessary for as.numeric() to work correctly
data$q5_eortc_fatigue_r1 = 
  as.numeric(str_replace(data$q5_eortc_fatigue_r1, ",", "."))  

sum(is.na(data$q5_eortc_fatigue_r1))  # 6 NAs introduced where cells were "!NULL"

# ----------------------------------- #
# For several variables it could be appropriate to convert them to factors but, at this stage of the analysis its rather optional.

# Health insurance
data$fc_caisse = factor(data$fc_caisse, labels = c("CNAMTS", "MSA", "RSI"))

# Pathology location
data$ms_codcancer = factor(data$ms_codcancer, 
                           labels = c("Breast", "Lung", "Colon & Rectum",
                                      "Prostate", "VADS", "Bladder", "Kidney",
                                      "Thyroid", "Lymphoma", "Melanoma", "Cervix",
                                      "Uterus"))

# Marital status
data$q5_sd4_r1 = factor(data$q5_sd4_r1, 
                        labels = c("Married/Partnered/Concubine", 
                                   "Single/Divorced/Separated/Widowed"))

# Children yes/non
data$q5_sd5 = factor(data$q5_sd5, labels = c("Yes", "Non"))

# Level of study
data$q5_sd10_r2 <- factor(data$q5_sd10_r2, 
                          labels = c("No degree", "Less than Bachelor's degree",
                                     "High school diploma or more"))

# Pain
data$q5_pain <- factor(data$q5_pain, labels = c("Yes", "Non"))

# Anxiety
data$q5_anxiete <- factor(data$q5_anxiete, 
                          labels = c("No anxiety", "Questionable anxiety state",
                                     "Certain anxiety state"))

# Depression
data$q5_depression <- factor(data$q5_depression,
                             labels = c("No depression", 
                                        "Questionable depression state",
                                        "Certain depression state"))

# Net salary category
data$q5_jobv5.36_r1 <- factor(data$q5_jobv5.36_r1, 
                              labels = c("<1500€", ">=1500€","Not employed"))

# Social category
data$ms_csp_enq_3c_r1 <- factor(data$ms_csp_enq_3c_r1, 
                                labels = c("Executives", "Managerial occupations", "Not employed"))

# Sequels
data$q5_med23.1 <- factor(data$q5_med23.1, 
                          labels = c("YES and they are very important", 
                                     "YES and they are important", 
                                     "YES but moderate", "YES but very moderate",
                                     "NO, i have no after-effects"))
```

```{r}
# Associationg a label with each variable, purely for QoL
label(data$fc_caisse) <- "Health insurance"
label(data$ms_codcancer) <- "Pathology's location"
label(data$fc_agediag_r0) <- "Age"
label(data$q5_sd4_r1) <- "Marital status"
label(data$q5_sd5) <- "Children yes/non"
label(data$q5_sd10_r2) <- "Level of study"
label(data$q5_pcs12_r1) <- "Physical QoL"
label(data$q5_mcs12_r1) <- "Mental QoL"
label(data$q5_pain) <- "Pain"
label(data$q5_anxiete) <- "Anxiety"
label(data$q5_depression) <- "Depression"
label(data$q5_jobv5.36_r1) <- "Net salary"
label(data$ms_csp_enq_3c_r1) <- "Socio-professional category"
label(data$q5_med23.1) <- "Sequels"
label(data$q5_eortc_fatigue_r1) <- "EORTC fatigue scale"
```

## 7. Definition of clinically significant fatigue score: score \>= 40 on the fatigue scale included in the survey, the threshold at which a fatigue condition was shown to be clinically significant. Create a categorical variable based on this definition.

```{r}
data$q5_eortc_fatigue_r1_fac = 
  cut(data$q5_eortc_fatigue_r1, 
      breaks = c(0,40,100), 
      labels = c("Not Clinically significant", "Clinically significant"), 
      include.lowest = T)

label(data$q5_eortc_fatigue_r1_fac) <- "EORTC fatigue scale"
```

## 8. Group the modalities of the variable sequelae into 3 modalities. This new variable, named "Q5_med23.1_rec" will be considered in the following analyses instead of "Q5_med23.1".

```{r}
data$q5_med23.1_rec = factor(data$q5_med23.1,
                             labels = c("Important sequelae","Important sequelae",
                                        "Moderate sequelae", "Moderate sequelae",
                                        "No sequelae"))

label(data$q5_med23.1_rec) = "Sequels"
```

## 9. Display the frequency table for this new variable.

```{r}
a = table(data$q5_med23.1_rec)
b = paste(round(prop.table(table(data$q5_med23.1_rec))*100,2),"%")

print(rbind(a,b))
```

## 10. Concerning age: What is the average age of our study population, then that of breast cancer.

```{r}
whole_pop = data$fc_agediag_r0
breast_pop = data$fc_agediag_r0[which(data$ms_codcancer=="Breast")]

mean(whole_pop) # Mean age of our population

mean(breast_pop, na.rm=T) # mean age for breast cancer subpopulation. 
```

### Determine the 95% confidence intervals (CI) for each of the calculated means.

```{r}
t.test(whole_pop)$"conf.int" # T confidence intervals for the whole population

t.test(breast_pop)$"conf.int" # T confidence interval for the breast cancer subpopulation
```

### Calculate the variance, standard deviation of the sample, then that of breast cancer.

```{r}
prettyR::describe(whole_pop,num.desc=c("var","sd"), 
                  xname="the age variable for the whole population",
                  horizontal=FALSE)

prettyR::describe(breast_pop,num.desc=c("var","sd"), 
                  xname="the age variable for the breast cancer population",
                  horizontal=FALSE)
```

## 11. Draw a graph that will represent the distribution of age by location of the pathology. Choose the most appropriate graph. Export the graph in a pdf format.

```{r}
# In order of personal preference
# Solution 1
ggplot(data = data, 
       aes(x=fc_agediag_r0, group = ms_codcancer, fill = ms_codcancer)) +
    geom_density()+
    scale_fill_brewer(palette="Set3")+
    scale_y_continuous(breaks = seq(0,0.1, 0.01), minor_breaks = seq(0, 0.005, 0.01))+
    scale_x_continuous(breaks = seq(0,100, 5))+
    labs(x = "Age",
         y = "Distribution probability",
         title = "Age distribution by cancer location",
         fill = "Cancer location")+
    facet_grid(vars(ms_codcancer))+
    theme(strip.text.y = element_text(size = 7))

ggsave("Solution 1.pdf", plot = last_plot(), device = "pdf", dpi = 300, width = 20, height = 35, units = "cm")

# Solution 2
ggplot(data = data, 
       aes(x=fc_agediag_r0, group = ms_codcancer, fill = ms_codcancer)) +
  geom_density()+
  scale_fill_brewer(palette="Set3")+
  scale_y_continuous(breaks = seq(0,0.1, 0.01), minor_breaks = seq(0, 0.005, 0.01))+
  scale_x_continuous(breaks = seq(0,100, 5))+
  labs(x = "Age",
       y = "Distribution probability",
       title = "Age distribution by cancer location",
       fill = "Cancer location")+
    facet_wrap(vars(ms_codcancer))

ggsave("Solution 2.pdf", plot = last_plot(), device = "pdf", dpi = 300, width = 40, height = 30, units = "cm")

# Solution 3
ggplot(data = data, 
       aes(x=fc_agediag_r0, group = ms_codcancer, colour = ms_codcancer)) +
    geom_density(stat = "bin", size = 1)+
    scale_color_manual(values = primary.colors(n=12, step = 6))+
    scale_x_continuous(breaks = seq(0,100, 5))+
    scale_y_continuous(breaks = seq(0,1000, 10))+
    labs(x = "Age",
         y = "Count",
         title = "Age distribution by cancer location",
         colour = "Cancer location")+ theme_excel()

ggsave("Solution 3.pdf", plot = last_plot(), device = "pdf", dpi = 300, width = 25, height = 20, units = "cm")
```

## 12. Determine the factors associated with physical and then mental quality of life, including variables with a p-value \< 0.2. Which model will you use? How will you proceed? Interpret the final result.

```{r}
# Listing the explicative variables
explicative_variables = c("fc_caisse", "ms_codcancer","fc_agediag_r0","q5_sd4_r1","q5_sd5","q5_sd10_r2","q5_eortc_fatigue_r1_fac", "q5_anxiete","q5_depression","q5_jobv5.36_r1","ms_csp_enq_3c_r1", "q5_med23.1_rec","q5_pain")

# Printing the list of explicative variables, with their labels, that are gonna be used in the model
kable(label(data[explicative_variables]))
```

```{r recoding reference mod for factor variables}
# Recoding the reference for qualitative variables
# List of factors to be recoded to make sure we correctly interpret the results
subset(sapply(data,is.factor), sapply(data,is.factor)==1)

data$fc_caisse <- relevel(data$fc_caisse, ref = "CNAMTS")
data$ms_codcancer <- relevel(data$ms_codcancer, ref = "Breast")
data$q5_sd4_r1 <- relevel(data$q5_sd4_r1, ref = "Married/Partnered/Concubine")
data$q5_sd5 <- relevel(data$q5_sd5, ref = "Non")
data$q5_sd10_r2 <- relevel(data$q5_sd10_r2, ref = "No degree")
data$q5_anxiete <- relevel(data$q5_anxiete, ref = "No anxiety")
data$q5_jobv5.36_r1 <- relevel(data$q5_jobv5.36_r1, ref = "Not employed")
data$ms_csp_enq_3c_r1 <- relevel(data$ms_csp_enq_3c_r1, ref = "Not employed")
data$q5_med23.1_rec <- relevel(data$q5_med23.1_rec, ref = "No sequelae")
data$q5_pain <- relevel(data$q5_pain, ref = "Non")
data$q5_depression <- relevel(data$q5_depression, ref = "No depression")
data$q5_eortc_fatigue_r1_fac <- relevel(data$q5_eortc_fatigue_r1_fac, ref = "Not Clinically significant")

# Checking if any of the factor levels has a low effective
data %>% select(where(is.factor)) %>% apply(2,table)
```

```{r Explicative model - linear regression}
# Physical QoL
mod1 <- lm(data = data, q5_pcs12_r1 ~ fc_caisse + ms_codcancer + fc_agediag_r0 + q5_sd4_r1 + q5_sd5 + q5_sd10_r2 + q5_eortc_fatigue_r1_fac + q5_anxiete + q5_depression + q5_jobv5.36_r1 + q5_med23.1_rec + q5_pain + q5_mcs12_r1)

# Mental QoL
mod2 <- lm(data = data, q5_mcs12_r1 ~ fc_caisse + ms_codcancer + fc_agediag_r0 + q5_sd4_r1 + q5_sd5 + q5_sd10_r2 + q5_eortc_fatigue_r1_fac + q5_anxiete + q5_depression + q5_jobv5.36_r1 + q5_med23.1_rec + q5_pain + q5_pcs12_r1)

# Factors associated to physical QoL score
subset(summary(mod1)$coefficients, 
       summary(mod1)$coefficients[, 4] < 0.2) %>% # filter for p-value<0.2
  data.frame() %>%
  arrange(Estimate) %>% # ordering according to the column Estimate
  kable(caption = "Factors associated to physical QoL score",
        col.names = c("Estimate", "Std error", "t-value", "P-value"))

# Factors associated to mental QoL score
subset(
  summary(mod2)$coefficients,
  summary(mod2)$coefficients[, 4] < 0.2) %>% # filter for p-value<0.2
  data.frame() %>%
  arrange(Estimate) %>% # ordering according to the column Estimate
  kable(caption = "Factors associated to mental QoL score",
        col.names = c("Estimate", "Std error", "t-value", "P-value"))

```

#### How to interpret the linear regression model

Printed tables have already been filter to exclude results with a p-value \>= 0.2 criteria.

Nota bene: these are *associations*, not implying causality. \
The huge population allow us to have great statistical power, lots of associations come out as statistically relevant but the magnitude of the effect is very small and frankly irrelevant on a scale that goes from 0 to 100 (or so it seems, we've no additional information on scale boundaries).

**Categorical variables**: under the column "Estimate", the table shows the average difference between a given modality and the reference modality for the same factor variable, as defined in the code chunk "*recoding reference mod for factor variables*".\
For example, patients in a "*Certain depression state*" score, on average, 5.63 points lower for Physical QoL compared to "*non-depressed*" patients.\
According to the same logic, someone with a monthly salary over 1500€ scores on average 2.78 points higher compaired to un unemployed patient.\
The same logic applies all others categorical variables and modalities.

**Quantitative variables**: two are the numeric variables taken into account by the model, patient's age and "the other" QoL score.\
In a plot where x="Age" and Y=QoL score, *Estimate* is the coefficient that ties the two variables.\
For example, for every additional year of age at diagnosis, physical QoL score lowers, on average, by 0.084 .

------------------------------------------------------------------------

**Key take-away messages:**

1)  *Depressed* and *anxious* patients score lower on both scales with an impact proportional to the severity of the psychiatric pathology.\

    Patients with *sequels* and *clinically significant fatigue* are also associated with scores notably lower.\

    [These 4 variables seems to be the only truly impact-full ones.\
    ]{.underline}

2)  Educational degree and monthly salary seems to have a modest impact only on the physical QoL score.

3)  Mental and physical QoL scores look to be inversely proportional but the coefficient is rather small, requiring huge variations on a scale to impact the other one. This is unexpected nevetheless, if both scales go in the same direction ( 0 \--\> 100).

4)  Age at diagnosis seems to have only a very small impact on both score, especially the mental one.

5)  A few cancer locations come out as statistically significant. As :

-   we have no additional information on the kind of therapy the patients underwent to or on the stade of their disease at diagnosis

-   coefficients are very small (the most important one being lung cancer patients scoring 2 points lower for physical QoL, on average)

-   results are inconsistent between the two score (lung cancer scoring the lowest for physical QoL and the highest for mental QoL score)

    no meaningful hypothesis/explanation can be formulated.

## 13. Export the new database in ".csv" format.

```{r}
# Writing the .csv in the working directory
write.csv(data,"new_BDD_VICAN_FM.csv", row.names = FALSE)
```

## 
